Complementing theory on adolescent emotion regulation using  machine learning

### 2022-03-14 manuscript/manuscript.Rmd
update 2022-03-14 08:25:05

authornote: |
  CVL conceived of the study, conducted the final analyses, and wrote the draft paper. LB conducted preliminary analyses and gave feedback on the draft. SB and WM provided data and gave feedback on the draft. This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author. Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp). RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University. The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript. If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.

abstract: |
  This paper set out to identify the most important early early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents with similar trajectories, based on their scores on the predictors. This person-centered approach not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories. The results indicated that individual differences (e.g., in personality) and conflict behaviors in relationships with parents and friends, and internalizing and externalizing were important early predictors. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

The present study seeks to overcome these limitations
by using data-driven insights to complement and reflect on existing theory.
To this end, a broad range of potential risk factors were analyzed using machine learning,
and ranked in terms of predictive importance.
<!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
Specifically,
the SEM-forests algorithm was used to
identify the most important early indicators of developmental trajectories
of emotion regulation across adolescence.
The resulting insights indicate which factors are most predictive of adolescents' diverging destinies in terms of emotion regulation development.
Armed with this knowledge, future research and interventions can focus on the most important predictors.
Moreover, these insights can guide the formation of overarching theory that adequately represents essential phenomena.
<!-- complement existing theory on adolescent emotion regulation using data-driven insights from machine learning. -->

@@

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014].
<!-- exposes youth to a risk of emotion dysregulation, -->
For example, they might not yet be sufficiently equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

@@

<!-- Maybe review theory briefly? -->
The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
A recent literature review did so
using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
Typically, phenomena are identified using narrative literature reviews.
However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008].
Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021]. <!-- reflection and meaning-making-->

@@

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to not only complete the cycle, but to enable confirmatory prediction testing growing from exploratory approaches.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

From wave 2-6, adolescents completed the difficulty in emotion regulation scale [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2018],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Score reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
To this end, we imposed a five-factor CFA (one for each wave) on the indicators of the dependent variable.
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the overall trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
As score reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

@@

The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
Consequently, we used single imputation using the missForest algorithm,
a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
This method is valid when missingness is either random or contingent on observed data (MAR or MCAR).
It does not make any distributional assumptions and does not extrapolate,
meaning it returns realistic values within the range of the observed data.
The proportion of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
<!-- Original publication had up to 30% missing, Wolputte, E. V., & Blockeel, H. (2020, October). Missing value imputation with MERCS: a faster alternative to MissForest. In International Conference on Discovery Science (pp. 502-516). Springer, Cham. looked at higher numbers. According to Mplus: minimum 10% -->

@@

As we set out to identifying important predictors of emotion regulation development,
this latent class analysis is not sufficient to answer our research question.
In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories?"
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories?"
This method is suitable to meet the study goals.

@@

**Note that a number of predictors were reverse coded; re-computing the analyses that generate the Figures requires access to a supercomputer, which we have not been able to obtain at this point. These reverse-coded plots have been highlighted in red. Reverse the + and - 1SD lines to interpret these.**

@@

```{r figpdp1, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 1-12. Red highlights indicate reverse-coded predictors."}
knitr::include_graphics("../plots/pdp1-12r.png")
```
```{r figpdp2, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 13-24. Red highlights indicate reverse-coded predictors."}
knitr::include_graphics("../plots/pdp13-24r.png")
```
```{r figpdp3, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 25-36. Red highlights indicate reverse-coded predictors."}
knitr::include_graphics("../plots/pdp25-36r.png")
```
```{r figpdp4, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 37-44. Red highlights indicate reverse-coded predictors."}
knitr::include_graphics("../plots/pdp37-44r.png")
```

@@

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous  sample from the Netherlands.
This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to enable easy replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.

>>>>

authornote: |
  This is a preprint paper, generated from Git Commit # `r gert::git_commit_id()`. CVL conceived of the study, conducted the final analyses, and wrote the draft paper. LB conducted preliminary analyses and gave feedback on the draft. SB and WM provided data and gave feedback on the draft. This work was funded by a NWO Veni Grant (NWO Grant Number VI.Veni.191G.090), awarded to the lead author. Data of the RADAR (Research on Adolescent Development And Relationships) study were used (https://doi.org/10.17026/dans-zrb-v5wp). RADAR has been financially supported by main grants from the Netherlands Organisation for Scientific Research (GB-MAGW 480-03-005, GB-MAGW 480-08-006), Stichting Achmea Slachtoffer en Samenleving (SASS), the Netherlands Organisation for Scientific Research to the Consortium Individual Development (CID; 024.001.003), a grant of the European Research Council (ERC-2017-CoG - 773023 INTRANSITION), and various other grants from the Netherlands Organisation for Scientific Research, VU University Amsterdam, and Utrecht University. The funders had no role in the study design, data collection and analysis, decision to publish, or preparation of the manuscript. If the regulations of the journal allow it, the participating children, parents and schools are to be thanked as well.

abstract: |
  This paper set out to identify the most important early early predictors of emotion regulation development from many (87) candidates using the machine learning algorithm SEM forests. Candidate predictors were selected based on a prior text mining systematic review of 6305 papers. Participants were 497 Dutch adolescents. Predictors were assessed at age 13, and emotion regulation development from age 14-18 was described using a quadratic latent trajectory model. SEM-forests were used to group adolescents based on their scores on the predictors, while maximizing the similarity of developmental trajectories within each group. Compared to other person-centered approaches, this method not only identifies groups of adolescents with similar trajectories, but also indicates why some adolescents follow similar trajectories. The results indicated that the most important early predictors (at age 13) were individual differences (e.g., in personality), conflict behaviors towards parents and friends, and internalizing and externalizing behavior. Less important were parenting behaviors, except autonomy support and invasive parenting, bullying and victimization, delinquent peers, and substance use. The implications of these findings for theory formation and intervention are discussed, and we present an open source risk assessment tool, ERRATA.

@@

The present study seeks to overcome these limitations
by using data-driven insights to complement and reflect on existing theory.
To this end, a broad range of potential risk factors were analyzed using machine learning,
and ranked in terms of predictive importance.
<!-- These predictors may include both risk factors that render some youth susceptible to emotion regulation difficulties, and correlates that can be used to identify at-risk youth. -->
Specifically,
the SEM-forests algorithm was used to
identify the most important early indicators, at age 13, of developmental trajectories
of emotion regulation across adolescence (age 14 through 18).
The resulting insights indicate which factors are most predictive of adolescents' diverging destinies in terms of emotion regulation development.
Armed with this knowledge, future research and interventions can focus on the most important predictors.
Moreover, these insights can guide the formation of overarching theory that adequately represents essential phenomena.
<!-- complement existing theory on adolescent emotion regulation using data-driven insights from machine learning. -->

@@

Emotion regulation,
defined as the ability to manage emotional experience and expression,
is critical for mental health [@aldaoEmotionregulationStrategiesPsychopathology2010, @schaferEmotionRegulationStrategies2017, @braetEmotionRegulationChildren2014]
and social competence [@zimmermannEmotionRegulationEarly2014] throughout life.
There is mounting evidence, however, that adolescents do not uniformly become better at regulating emotions: 
staggered development of emotional and regulatory brain circuits gives rise to a "maturity gap" in mid-adolescence [@croneUnderstandingAdolescencePeriod2012].
The emotional salience of social stimuli is intensified, while the ability to regulate these emotions lags behind [@caseyBrakingAcceleratingAdolescent2011, @vanlissaLongitudinalInterplayAffective2014].
This imbalance motivates most adolescents to engage in adaptive exploration:
Seeking out experiences required to develop mature socio-emotional competences.
At the same time, this imbalance renders some adolescents vulnerable to emotion dysregulation [@leeAdolescentMentalHealth2014],
<!-- exposes youth to a risk of emotion dysregulation, -->
for example, because they are not yet well-equipped to regulate the pursuit of unhealthy incentives,
or persevere in pursuing adaptive goals when facing obstacles.
Thus, some become trapped in a negative spiral characterized by more negative and volatile emotions [@craccoEmotionRegulationChildhood2017],
which has been linked to the emergence of several mental disorders [@schaferEmotionRegulationStrategies2017].
<!-- This imbalance motivates most youngsters to seek out new challenges in life and love, -->
<!-- but renders some of them vulnerable to emotion dysregulation1,10. Unfortunately, there is little theory about which risk factors and environmental hazards render adolescents susceptible to emotional difficulties11. -->
<!-- Acquiring adaptive emotion regulation skills in adolescence plays a key role in this regard7,20, because t -->

@@

<!-- Maybe review theory briefly? -->
The first step towards theory formation or revision is identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020].
<!--Thus, offering a starting point for this endeavor,<!-- empirical investigations of predictors of adolescents' emotion regulation development -->
A recent literature review did so
using a text mining systematic review of author keywords and abstracts of 6305 publications on  adolescents’ emotion regulation [@vanlissaMappingPhenomenaRelevant2021].
Typically, phenomena are identified using narrative literature reviews.
However, these are often limited to a small number of papers and structured around well-established ideas [@littellEvidencebasedBiasedQuality2008]
Text mining systematic reviews, by comparison, are relatively comprehensive and derive insights from the literature through a relatively transparent and objective - albeit less interpretative - procedure [@vanlissaMappingPhenomenaRelevant2021]. <!-- reflection and meaning-making-->

@@

According to models of knowledge production within the quantitative positivist paradigm,
a field progresses through the cyclical alternation of exploratory and confirmatory research, also known as the "empirical cycle" [@degrootMethodologieGrondslagenVan1961, @wagenmakersCreativityVerificationCyclePsychological2018].
Exploratory research gives rise to testable predictions,
and confirmatory research evaluates testable predictions on new data.
Crucially, these phases are linked through evaluation of existing theory in light of empirical findings [@degrootMethodologieGrondslagenVan1961].
The fact that our field still lacks overarching theory despite decades of productive empirical research
suggests that the empirical cycle is not routinely completed.
In support of this claim, it has been noted that the overwhelming majority of research in our field appears, at first glance, to be confirmatory [@demuthQualitativeMethodologyDevelopmental2015].
More exploratory research is thus needed to complete the cycle.
Secondly, and paradoxically given the first argument,
the link between confirmatory research and theory is often tenuous.
For instance, many studies derive their predictions from prior empirical research, not from theory [e.g., @vanlissaCostEmpathyParentadolescent2017].
Thirdly, theories are rarely falsified or revised in light of empirical findings.
This failure to evaluate theory leads to flexible, "invulnerable" frameworks that can accommodate any result [@szollosiArrestedTheoryDevelopment2021].
To advance theory formation, our field might benefit from "closing the empirical cycle" through increased exploratory research and reflection on theory.

@@

From wave 2-6, adolescents completed the difficulty in emotion regulation scale [DERS, @gratzMultidimensionalAssessmentEmotion2004].
Based on prior psychometric analyses [see @vanlissaRoleFathersMothers2018],
we used 24 items of this 32-item scale
(e.g., "When I’m upset, I have difficulty thinking
about anything else").
Responses ranged
from 1 (Almost never) to 5 (Almost always).
Reliability was excellent in all waves, $\alpha \in [`r report(min(alphs), equals = F)`, `r report(max(alphs), equals = F)`]$.
We used latent variable scores (instead of mean scores) for further analyses.
To this end, we imposed a five-factor CFA (one for each wave) on the indicators of the dependent variable.
Latent means were restricted to zero, such that resulting scores reflect the deviation of individual adolescents' trajectories from the overall trajectory.
As measurement invariance must be imposed in order to compare latent scores across waves,
we compared a model with configural invariance ($BIC = 141247.46$)
to a model that imposed metric invariance ($BIC = 140913.27$).
The lower BIC indicates support for metric invariance.
The RMSEA and SRMR of the invariant model indicated acceptable fit ($\text{RMSEA} = .05, \text{SRMR} = 0.05$), but the CFI indicated poor fit ($\text{CFI} = .80$).
As reliability and standardized factor loadings were high across all waves ($\lambda \in [0.44, 0.84]$),
we considered it acceptable to proceed with individual factor scores for further analysis.

@@

The SEM-forests algorithm requires complete data and cannot accommodate multiple imputed data.
Consequently, we used single imputation using the missForest algorithm,
a random forests based approach with comparable performance to multiple imputation and FIML estimation [@stekhovenMissForestNonparametricMissing2012].
This method is valid when missingness is either random or contingent on observed data (MAR or MCAR).
It does not make any distributional assumptions and does not extrapolate,
meaning it returns realistic values within the range of the observed data.
The percentage of missingness ranged from $`r report(min((495-desc_cat[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for categorical data (see Table \@ref(tab:tabcat)),
and from $`r report(min((495-desc_num[["n"]])/495), equals = F)` - `r report(max((495-desc_cat[["n"]])/495), equals = F)`$ for continuous data (see Table \@ref(tab:tabnum)).
<!-- Original publication had up to 30% missing, Wolputte, E. V., & Blockeel, H. (2020, October). Missing value imputation with MERCS: a faster alternative to MissForest. In International Conference on Discovery Science (pp. 502-516). Springer, Cham. looked at higher numbers. According to Mplus: minimum 10% -->

@@

As we set out to identifying important predictors of emotion regulation development,
this latent class analysis is not sufficient to answer our research question.
In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories?"
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories"?
This method is suitable to meet the study goals.

@@

@@

```{r figpdp1, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 1-20."}
knitr::include_graphics("../plots/pdp1-20.png")
```
```{r figpdp2, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 21-40."}
knitr::include_graphics("../plots/pdp21-40.png")
```
```{r figpdp3, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 41-60."}
knitr::include_graphics("../plots/pdp41-60.png")
```
```{r figpdp4, fig.cap="Predicted trajectory of difficulties in emotion regulation for predictors 61-87."}
knitr::include_graphics("../plots/pdp61-87.png")
```

@@

The study also has limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogeneous  sample from the Netherlands.
This raises specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is made available in a reproducible archive to enable easy replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
It is also important to consider that, in our efforts to identify important early predictors of emotion dysregulation, we relied on predictors at a single timepoint (age 13).
It is possible that the relative importance of these predictors changes over time,
or even dynamically, in interplay with emotion regulation.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.
